---
title: "Lab 6: Sports"
date: "10/05/22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Part I: Understanding the Context of the Data

1. Please write here the sport you are focusing on.

2. Write out two descriptive and two predictive questions that we could answer your sport of choice using statistics.

---

You will find that your sport has two data-sets associated with it: a *teams* dataset and a *players* dataset. You can find excerpts of each dataset in a file attached to Ed. Use these excerpts to answer the following questions.

3. What is the unit of observation for the *teams* data set? What about for the *players* data set?

4. What is a question you could answer using the *teams* data set but not the *batting* data set, and vice versa?

5. What is a question that we would need more granular (measured on a finer/more specific part of the game) data than the *teams* and *players* data set provide to answer?

6. Choose a variable from your dataset. Based on what values you expect it to take and its place in the Taxonomy of Data:

-   sketch a visualization of this variable. label and title your axes, and provide the aesthetic mapping and the geometry you are using. 
-   In one or two sentences, describe your visualization in terms of shape, spread and center, and explain these choices.


7. Some people believe analytics is ruining sports because teams are more cautious which makes the games less entertaining. Do you agree or disagree? Why?

## Part II: Computing on the Data

Use the following code to load your data-sets into the environment pane from the `stat20data` library. The data are called *SPORT_teams* and *SPORT_players*, respectively. Here, we load the two datasets for football.

```{r}
library(stat20data)
data(football_teams)
data(football_players)
```

The following questions apply to the *teams* datasets. 

8. Subset the *teams* dataset to only include years from 2000 to present day. What are the dimensions of this filtered dataset?

11. Plot the relationship between **IMPORTANT METRIC HERE** and wins. Describe the relationship. How does it compare to the relationship between runs and wins?

12. Fit a simple linear model to predict **IMPORTANT METRIC HERE** by runs. Write out the equation for the linear model report the $R^2$.

13. What is the average number of season **IMPORTANT METRIC HERE** and wins? Based on the previous model, how many games would you predict a team that scored the average number of runs would win? What about a team that scored 600 runs? What about 850 runs?

14. Fit a multiple linear regression model to predict wins by **IMPORTANT METRIC 1 HERE** and **IMPORTANT METRIC 2 HERE** allowed. Write out the equation for the linear model and report the $R^2$. How does this model compare to the simple linear regression from the previous question?

15. Fit a multiple regression model to predict **IMPORTANT METRIC 1 HERE** using at least two other variables in this data set. How does the $R^2$ value change? Do you think the new model you created predicts wins better?

The following questions apply to the *players* datasets. 

TBD

