---
title: "Introducing Probability"
subtitle: "PS"
format: stat20handout-pdf
---

Consider the seven sided die (with sides 0, 1, 2, 3, 4, 5 and 6) from your reading questions. You roll this die once. Let:

-   A be the event that a given roll yields an even number.   
-   B be the event that a given roll is greater than or equal to three.   
-   C be the event that the number appears in the phrase “Stat 20”.

@. Find $\text{P}(A)$, $\text{P}(B)$ and $\text{P}(C)$

  \
  
  \
  
@. Find $\text{P}(A \cup B)$. 

  \
  
  \
  
@. Find $\text{P}(B \cap C)$

  \
  
  \
  
@. Find $\text{P}(A \cap B^C)$.

  \
  
  \



One ticket will be drawn at random from each of the two boxes below:

\begin{center}
$A: \fbox{\fbox{1} \fbox{2} \fbox{3}}$ $\hspace{1cm}$ $B: \fbox{ \fbox{1} \fbox{2} \fbox{3} \fbox{4}}$
\end{center}

@. What is the probability the number drawn from $A$ is greater than the one drawn from $B$?

   \
   
   \

@. What is the probability that the number drawn from $A$ is equal to the one drawn from $B$?

   \
   
   \

@. What is the probability the number drawn from $A$ is smaller than the one drawn from $B$?

{{< pagebreak >}}
   

Consider an outcome space $\Omega$ with events $A, B$ that are *not* mutually exclusive. Let $P(A) = 0.5, P(B) = 0.7$ and $P(A \cap B) = .4$.

@. Draw a Venn Diagram containing two circles, one representing $A$ and the other $B$.

  \
  
  \
  
  \

@. Shade in the space in the diagram above corresponding to event $A$.

@. Shade in the space in the diagram above corresponding to event $B$ in a way that is different than the way you shaded in the space for event $A$.

@. Based on your shading, calculate the probability $P(A \cup B)$. Explain why you cannot add $P(A)$ and $P(B)$ directly to obtain this answer.  

  \
  
  \

Return to the example of the seven-sided die from your reading questions and earlier in this problem set. Write code in R (and write the code below) to:

@. Simulate rolling the seven-sided die once.

  \
  
@. Simulate rolling the seven-sided die seven times.

  \
  
@. Render a Quarto document with parts 1 and 2 such that you receive the same sample each time you run the code. You can just provide the line of code which is responsible for this behavior.

  \
  
Examine the following code.

```{r}
#| eval: false
# First create a vector with the possible sums of a pair of dice
dice <- seq(from = 2, by = 1, to = 12)
# Then sample once from these sums 
sample(x = dice, size = 1)
```


@. If we want to simulate rolling a pair of dice and summing the spots (for say, playing Monopoly), will the following code do this correctly? Explain.


\

@. If you said that the code is incorrect, modify the following code to simulate this process correctly. *Hint: you may want to use `sample()` more than once.*



<!--
@. If we want to simulate rolling a pair of dice and summing the spots (for say, playing Monopoly), will the following code do this correctly? Explain.

```{r}
#| eval: false
# first create a vector with the possible sums of a pair of dice
dice <- c(2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
# then sample once from dice
sample(dice, 1)

```

\ 

\

\


@. If we are defining a box to simulate tossing a fair coin three times and counting the **number** of heads, would either of the methods below work? If not, why not? (Note that, for example, the box  $\fbox{\fbox{0} \fbox{1}}$  would be represented by the `c(0, 1)` in R, and we would then use the function `sample()`.

(a) $\fbox{\fbox{0} \fbox{1}}$ - Draw three times at random with replacement, and sum the draws.
(b) $\fbox{\fbox{0} \fbox{1} \fbox{2} \fbox{3}}$ -  Draw once, the result is the number of heads.


\

\

\

<!--
@. I want to estimate the proportion of people in Berkeley who speak at least two languages. I stand at the corner of University and Shattuck and ask each person who goes by how many languages they speak, and keep a count of how many speak at least two. Is this a good way to estimate the proportion I am looking for? 

   \

   \
-->


<!--
(a) Consider the box $\fbox{ \fbox{1} \fbox{2} \fbox{2} \fbox{3} \fbox{4}}$. I draw two tickets at random **with** replacement. If my first draw is a 2, what is the probability that my second draw is a 3?

\

\

(b) Consider the box $\fbox{ \fbox{1} \fbox{2} \fbox{2} \fbox{3} \fbox{4}}$. I draw two tickets at random **without** replacement. If my first draw is a 2, what is the probability that my second draw is a 3?

\

\



@. Consider an outcome space $\Omega$ and events $A, B, C$ with $P(A) = 0.6$, $P(B) = 0.7, $.

(a) What is the smallest possible value of $P(A \cap B) = P(A \textbf{ and } B)$

\
\

(b) What is the largest possible value of $P(A \cap B)$?

\
\

(c) What is the smallest possible value of $P(A \cup B)$? 

\
\

(d) What is the largest possible value of $P(A \cup B)$? 

\
\

@. Think about the two types of plots you saw in the notes: the empirical distributions and the actual probability distribution for the outcomes of rolling a fair six-sided die. Suppose we play a game now: we roll a die $n$ times, and you get 5 dollars if the proportion of times you roll a 6 is greater than 20%. You get to choose $n$ to be either 10 or 1000. Which one would you choose? Why?
-->

 
