---
title: "Practice Problems"
subtitle: "Confidence Intervals"
format: pdf
---

:::notes
Goal: Give students practice calculating confidence intervals and talking through their interpretation and what effects their width

These practice problems can be done in class in pairs/trios or at home for students who aren't in class. They will become part of the next problem set. The slides introduce / practice several functions used here.

Suggesting timeline:

1. Review functions in slides
2. Make students work in at least pairs for the handout. Ask that they draw their confidence intervals for the mean on the front board. You'll want to pre-populate the axes carefully and probably have them draw their intervals vertically so that they will all fit. Draw one on the board to demonstrate.
3. At the end of the class period, review answers.
:::

## Part I: 

The goal of these practice problems is to experiment with the notion of a sampling distribution and confidence intervals. In any applied setting, you only have access to a sample of the data, but for these practice problems we'll work with a data set where we have access to the full population: the flights data set.

#### Question 1

Load the `flights` data from the `stat20data` package that you used in Lab 4.

```{r}
#| echo: true
library(tidyverse)
library(stat20data)
data(flights)
```

Then:

- Set a seed of your choosing
- Take a random sample of 100 flights from `flights`
- Calculate the mean and SD of the departure delays
- Assuming the the sampling distribution is roughly normal, what is your 68% confidence interval? Calculate it.

*SOLUTION*

```{r}
library(tidyverse)
library(stat20data)
data(flights)

set.seed(302)
stats <- flights %>%
  slice_sample(n = 100) %>%
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE),
            sd_delay  = sd(dep_delay, na.rm = TRUE))

c(stats$avg_delay - 1 * stats$sd_delay / sqrt(100), 
  stats$avg_delay + 1 * stats$sd_delay / sqrt(100))
```

#### Question 2

Did your interval contain the population mean? This is a rare case where you can check by calculating the average departure delay of the entire `flights` data set. If you repeated process 100 times - taking how many of the resulting intervals would you expect to contain the population mean?

*SOLUTION*

```{r}
flights %>%
  summarize(mean(dep_delay, na.rm = TRUE))
```

Yes, the interval contains the population parameter. You would expect around 68 of the intervals to contain the population mean.


#### Question 3

Select a second population parameter to estimate, but in this case choose a **proportion** instead of a mean. For example: the proportion of flights that departed ahead of schedule (don't use this example). Repeat the steps in question 1 to take a SRS of 100 flights and form a 68% confidence interval, adapting it for a proportion by using the appropriate modifications from the notes.

*EXAMPLE SOLUTION*

```{r}
set.seed(302)
stats <- flights %>%
  slice_sample(n = 100) %>%
  summarize(p_ahead = mean(dep_delay < 0, na.rm = TRUE),
            sd = sqrt(p_ahead * (1 - p_ahead)))

c(stats$p_ahead - 1.96 * stats$sd / sqrt(100), 
  stats$p_ahead + 1.96 * stats$sd / sqrt(100))
```


#### Question 4

Did your interval contain the population proportion? This is a rare case where you can check by calculating the actual proportion found in the entire `flights` data set. If you repeated process 100 times - taking how many of the resulting intervals would you expect to contain the population proportion?

*SOLUTION*

```{r}
flights %>%
  summarize(p_ahead = mean(dep_delay < 0, na.rm = TRUE))
```

Yes, the interval contained the true population parameter. You would expect 68 of the resulting intervals to contain the population proportion

#### Question 5

For both of the settings above - estimating a population mean and a population proportion - what are *two* things that can be changed to decrease the width of the confidence intervals?

*SOLUTION*
1. Decrease the confidence level
2. Increase the sample size
3. (Much harder to realize) Decrease the SD of the population

#### Question 6

In these practice problems we treated the `flights` data as the target population (and access frame) then took simple random samples of size 100. In reality, the entire `flights` data set was the sample of data that we investigated in Lab 4.

Give an example of a larger target population to which you would be comfortable generalizing from the full `flights` data set. Provide another example of one to which you would *not* be comfortable generalizing. It may be helpful to sketch scope diagrams.

*EXAMPLE SOLUTIONS*
- Comfortable generalizing certain parameter estimates to San Jose Airport.
- Comfortable generalizing from pre-pandemic to beforehand - late 2019.

- Uncomfortable generalizing to Taipei International Airport or other Airport that are significantly different.
- Uncomfortable generalizing to very different time spans.