---
title: "A Grammar of Graphics"
format: stat20slides-revealjs
---

## Agenda

- Concept Questions
- Lab 2: Class Survey
  - Part I
- *Break*
- Lab 2: Class Survey
  - Part II

# Concept Questions

##

:::{.poll}
What are the aesthetics and geometry of this plot?

Answer at `pollev.com`
:::

```{r}
#| fig-align: center
#| fig-width: 11
library(tidyverse)
library(palmerpenguins)

ggplot(penguins, aes(x = bill_length_mm,
                     fill = species)) +
    geom_density() +
    theme_gray(base_size = 20)
```

```{r}
countdown::countdown(1, top = 0)
```

##

:::{.poll}
What has changed when moving from the left to the right?
:::

```{r}
#| fig-align: center
#| fig-width: 11
library(patchwork)
p1 <- ggplot(penguins, aes(x = bill_length_mm,
                           y = bill_depth_mm,
                           color = species)) +
    geom_point(size=2) +
    theme_gray(base_size = 20) +
    theme(legend.position = "none")
p2 <- ggplot(penguins, aes(x = bill_length_mm,
                           y = bill_depth_mm,
                           shape = species)) +
    geom_point(size=2) +
    theme_gray(base_size = 20) +
    theme(legend.position = "none")
p1 + p2
```

```{r}
countdown::countdown(1, top = 0)
```


## 

You will be watching a 2.5 minute video of a presentation by a scientist, Hans Rosling, who studied global public health. He presents data visualizations depicting the change in life expectancy and family size over several decades in the 20th century.

:::{.poll}
On a piece of note paper:

- Sketch out the data frame used to create the graphic and add the names of the variables.
- List the aesthetic attributes used to encode the data in the graphic.
- Identify the geometry used in the plot.
:::

## {data-menu-title="Hans Rosling's Gapminder"}

<center>
<iframe src="https://embed.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen" width="854" height="480" style="position:absolute;left:0;top:0;width:100%;height:100%" frameborder="0" scrolling="no" allowfullscreen></iframe>
</center>

##

Please turn to your neighbors and...

:::{.poll}
Discuss what you came up with in terms of . . .

- the variables present in the data frame
- the aesthetic attributes used to encode that data in the plot
- the geometry
:::

```{r}
countdown::countdown(4, top = 0)
```


## What were the variables and aesthetic attributes?

. . .

:::: {.columns}

::: {.column width="50%"}

#### Visual Cues / Aesthetics

- Location along the x-axis
- Location along the y-axis
- Size of point
- Color of point
- Animation
:::

::: {.column width="50%"}
#### Variables

- Fertility rate
- Life expectancy
- Population
- Region
- Year
:::

::::


## What did the data frame look like?


What was the unit of observation? What were the variables? What were their type?

. . . 

:::: {.columns}

::: {.column width="50%"}
#### Unit of observation

- A country in a given year
:::

::: {.column width="50%"}
#### Variables

- Fertility rate (continuous)
- Life expectancy (continuous)
- Population (continuous)
- Region (nominal)
- Year (discrete)
:::

::::

```{r echo = FALSE, eval = FALSE}
## don't use the gapminder package data. Remake to have the original vars: life expectancy and fertility rate
library(gapminder)
gapminder %>%
  slice(1, 10) %>%
  kable() %>%
  kable_styling(latex_options = "striped")
```


## What geometry is used to represent the observations?

. . .

- Points

# Break

# Lab 2: Class Survey
