---
title: "Method of Least Squares"
format: stat20slides-revealjs
---

## Agenda

- Announcements
- Concept Questions
- Problem Set 16: *The Method of Least Squares*
- Lab 5.1: Understanding the Context of the Data

## Announcements

-   Problem Sets:
    -   PS 16 released Tuesday and due next Tuesday at 9am
    -   Extra Practice released Thursday (non-turn in)
  
. . . 

-   Lab 5:
    -   Lab 5.1 released Tuesday and due next Tuesday at 9am
    -   Lab 5.2 released Thursday and due next Tuesday at 9am
    -   Lab 5 Workshop next Monday

# Concept Questions

## Concept Question 1

An engineer working for Waymo self-driving cars is working to solve a problem. When it rains, reflections of other cars in puddles can disorient the self-driving car. Their team is working on a model to determine when the self-driving car is seeing a reflection of a car vs a real car.

:::{.poll}
Think of a potential response and predictor, and about whether this is a  regression or classification problem.
:::

```{r}
countdown::countdown(1)
```

:::{.notes}
This is a serious challenge encountered by self-driving cars at the moment. This is probably best thought of as a classification problem, with the response being either reflection / not or puddle / not. The predictors could be sensor input (cameras, lidar) as well as weather info (to know if it's been raining).
:::

## Concept Question 2

-   Here is a function `f`.

```{r}
#| echo: true
f <- function(x, y) {
  y*(x + 3) 
}
```

. . . 

:::{.poll}
What will the following line of code return?
:::

. . . 

```{r}
#| echo: true
#| eval: false
f(3,5)
```

:::{.notes}
This question reminds students that if function arguments are unnamed, they are assumed to have been input in the order they were originally specified as in the function definition. `x` will be assigned `3`, and `y` will be assigned `5`, so the correct answer is 30. 
:::

<!---
## Scenario 2

An analyst working for the UC Berkeley Admissions Office is working to help the university decide how many students to send offer letters to. They have a target class size (that fits within the budget and residence halls), but they're not sure how many students will accept the offer. How many should they admit?

:::{.poll}
Please identify A. the response, B. the predictor(s), and 
C. whether it is regression or classification (poll only has C.)
:::

```{r}
countdown::countdown(1)
```

:::{.notes}
This is another very real task. Universities use what's called a "yield model" to predict whether or not a student will accept the offer. That's a classification method. Those classifiers aggregated up across all offers lead to a numerical prediction for the total yield, which looks more like a regression problem, so either answer could work. It could also be a direct regression model where they use aggregate stats of the class to predict the total yield.

Response is either the decision of a single student or the total number of students that accept. Predictors could be either student level covariates (for classification) or class-level averages (for regression). Those might also include broader economic indicators like unemployment rate.
:::
-->


# Problem Set 

```{r}
countdown::countdown(25)
```


# Break

# Lab 

```{r}
countdown::countdown(25)
```

# End of Lecture
