---
title: "Practice Problems"
subtitle: "Regression"
---

```{r}
library(tidyverse)
library(stat20data)
data(flights)
```

#### Question 1

Create a scatter plot of the relationship between distance traveled and average speed in miles per hour of every flight that left the bay area in 2020.

```{r}
flights <- flights %>%
    mutate(avg_speed = distance / (air_time / 60))

flights %>%
    ggplot(aes(x = distance,
               y = avg_speed)) +
    geom_point()
```

#### Question 2

Use the `lm()` function to fit a linear model by least squares that predicts the average speed of the flight based on the distance and save it to `m1`. Recall the general form of arguments is `lm(y ~ x, data)`.

```{r}
m1 <- lm(avg_speed ~ distance, data = flights)
```

#### Question 3

What is the equation of the linear model that you have fit?

```{r}
m1
```

predicted-avg-speed = 351.4 + .078 \* distance

#### Question 4

What average speed would does your model predict for a flight travelling 0 miles? 500 miles? 5000 miles? Which of these, if any, do you expect to be accurate and why?

```{r}
newx <- data.frame(distance = c(0, 500, 5000))
predict(m1, newx)

351.4 + .078 * 5000
```

#### Question 5

Remake your scatter plot but add a line representing your linear model by adding a layer with `geom_smooth(method = "lm")`. How could you modify your model to improve the predictions that it makes?

```{r}
flights %>%
    filter(dest %in% c("HOU", "AUS", "DFW")) %>%
    ggplot(aes(x = distance,
               y = avg_speed)) +
    geom_point() +
    geom_text(aes(label = dest))
```

We could use a curved line instead of a straight one.
