---
title: "PS 9: Simulations"
---

```{r}
#| message: false
#| echo: false
#| warning: false

library(tidyverse)
library(patchwork)
```

In the following questions, use the chunks and uncomment the code by removing the # sign.

#### Question 1 Simulating the probabilities in de Méré's paradox, for 4 rolls of a die

- First we will create a vector that represents a die, and remember - the only outcome you are interested in is rolling a 6, so let's call it die6. Why does the box have 5 zeroes and 1 one?

```{r}
 die6 <- c(0,0,0,0,0,1)
```

- Now we want to sample from this box 4 times, to simulate rolling a die 4 times, and then count the number of sixes in these 4 rolls. Since the draws are 0's and 1's, if we sum the draws, that counts how many sixes we rolled. 

```{r}
 sum(sample(die6, size = 4, replace = TRUE))
```

- In the next code chunk, we will use the function `replicate()` to repeat the process of simulating 4 die rolls 100 times. 
- Each time we roll the die 4 times (using `sample()`, we will use `sum()` to counting the number of sixes in the four draws. So we are simulating rolling a die 4 times and counting the number of sixes, over and over, 100 times. 
- After we run the chunk below, we will have a vector `game1_100` that has 100 entries, each of which represents the number of sixes in 4 rolls.

```{r}
 game1_100 <- replicate(100, 
                  sum(sample(die6, size = 4, replace = TRUE)))
```



```{r}

```


Saywewanttodraw2ticketsfromtheboxandfindthechancethat
the 2nd ticket is 5:
1 2 3
box1 6 simulateroll
I 4
> box=1:6 ##but only want to count 5
ads
5
6
of
• How would we simulate this?
sum ample box 2 replacTe RUE areabottts
> box1=c(1, rep(0,5)) > sample(box1, 2)[2]
ble don't times megets wantto count of
00
##[]extracts values satisfying condition
> x=replicate(100000, sample(bOox1, 2)[2]) replicate N eerson ## usingDlong run definition of probability
> sum(x)/100000

