---
title: "Multiple Linear Regression"
format: stat20slides-revealjs
---

```{r}
#| message: false
library(tidyverse)
library(stat20data)
```

## Agenda

-   Announcements
-   Multiple Linear Regression Refresher
-   Quiz Review (this week's notes)
-   Break
-   Lab 2.2 (extended)

## Announcements

-   Quiz 1 is Monday, in-class and covers all lectures from the beginning to the semester until today.

. . . 

-   Lab 2.2, Problem Set 4 and Problem Set 5 are due Tuesday 9am
    -   Make sure you follow Lab Submission Guidelines on Ed

. . . 

-   RQ: Introducing Probability due on Monday/Tuesday at 11:59pm; Probability unit begins next week

. . . 

-   *Extra Practice* for Multiple Linear Regression added to the resources tab on the [course home page](https://stat20.org). 

# Multiple Linear Regression Refresher 

-   Head to `pollev.com` for a set of rapid-fire questions on last night's notes.


# Quiz Review

-   Head to `pollev.com` for a set of quiz-level questions pertaining to *Summarizing Numerical Associations* and *Multiple Linear Regression*.


## Question 1

```{r}
#| echo: true
m1 <- lm(bill_depth_mm ~ bill_length_mm, data = penguins)
```

. . . 

```{r}
#| echo: true
m2 <- lm(bill_depth_mm ~ bill_length_mm + body_mass_g + species, 
         data = penguins)
```

:::{.poll}
How many more coefficients does the second model have than the first?
:::

:::notes
Remind students that they need to remember whether `species` and `body_mass_g` are numerical or categorical. Students should know how many species there are (three). 

One addl. coefficient for body mass; two addl. for species (one less than the number of species, which is three). This gives a total of three more coefficients.
:::

# Questions 2-4

Consider the following multiple linear regression model, which will be the subject of the next three review questions.

## Question 2

```{r}
countdown::countdown(1, top = 0)
```

\

```{r}
#| echo: true
m2
```

:::poll
Which is the correct interpretation of the coefficient in front of **bill length**? *Select all that apply*.
:::

:::notes
This one assesses their ability to use a conditional interpretation of a regression coefficient (controlling for the other variables in the model...).

The correct interpretation controls for both other variables (body mass and species), and has the x variable and the y variable in the correct places. 
:::

## Question 3

```{r}
countdown::countdown(1, top = 0)
```

\

```{r}
#| echo: true
m2
```

:::poll
Which is the correct interpretation of the coefficient in front of **Gentoo**?
:::

:::notes
This one assesses their ability to use a conditional interpretation of a regression coefficient (controlling for the other variables in the model...)

The correct interpretation needs to involve a comparison to the reference level, Adelie and should not involve anything about bill length or body mass, (other than to say they should be fixed).
:::

## Question 4

```{r}
countdown::countdown(1, top = 0)
```

\

```{r}
#| echo: true
m2
```

:::poll
How would this linear model best be visualized?
:::

:::notes
Three numerical variables means we will have planes in a 3D space. The indicator coefficients will shift the planes up and down. There are three species, so three parallel planes is what we are looking for.
:::

## Question 5

Consider the following linear regression output where the variable `school` is categorical and the variable `hours_studied` is numerical.

| Coefficients     | Estimate |
| ----------- | ----------- |
| `(Intercept)`      | 2.5       |
| `hours_studied`        | .2        |
| `schoolCal`      | 1       |
| `schoolStanford`        | -1      |

## Question 5 (cont.)

-   Say I wanted to create a data frame from the original `edu` dataframe which contains the minimum, median, and IQR for `hours_studied` among each school. In order to do this, I make use of `group_by()` followed by `summarize()`. I save this data frame into an object called `GPA_summary`.

. . .

:::poll
What are the dimensions of `GPA_summary`?
:::

. . . 

```{r}
countdown::countdown(minutes = 1)
```

. . . 

:::notes
The correct answer in the poll should be 3x4. The three rows are for the three levels in the school category (there *is* an additional level beyond `"Cal"` and `"Stanford"`, regardless if it is not stated).  One column is for the school name, the other three are for each of the statistics calculated; the three rows are for the three levels in the school category.
:::

# Break

```{r}
countdown::countdown(minutes = 5, top = 0)
```

# Lab 2.2 (extended)

```{r}
countdown::countdown(minutes = 40, top = 0)
```

# End of Lecture

<!---

# Concept Questions

## Estimate the coefficient

```{r}
countdown::countdown(1, top = 0)
```

```{r}
library(tidyverse)
library(stat20data)
```

\
\

```{r}
#| echo: true
m1 <- lm(bill_depth_mm ~ bill_length_mm, data = penguins)
```

\
\
\

:::{.poll}
What will be the sign of the coefficient for `bill_length_mm`?
:::

:::notes
This one warms students up to the idea of working between the formula, the equation of the line, and a plot. It will also likely surprise them: without controlling for the species, there's actually a slight negative relationship here.
:::


## 

```{r}
#| echo: true
m1
```

. . .

```{r}
#| fig-align: center
penguins %>%
  ggplot(aes(x = bill_length_mm,
             y = bill_depth_mm)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()
```


## Estimate the coefficient, take two

```{r}
countdown::countdown(1, top = 0)
```

\

```{r}
#| echo: true
m2 <- lm(bill_depth_mm ~ bill_length_mm + species, penguins)
```

\
\
\

:::{.poll}
What will be the sign of the coefficient for `bill_length_mm`? How many coefficients will be in this linear model?
:::

:::notes
After the first question, some should get the second one correct, but the fact that it can change may surprise some students. They haven't need an example of a categorical variable with more than k = 2 levels, so it will be challenging for them to think through how many coefficients are required (since there are k = 3 levels, there are k = 3-1 dummy variables required, for a total of 4 coefficients.)

You may want to follow this question up by drawing on a board a data frame with `bill_length_mm` and `species` and show how you need to mutate two new dummy variables. You can also mention that by default, R leaves as the references level the one which comes first alphabetically.
:::

## 

```{r}
#| fig-align: center

penguins <- mutate(penguins, yhat = fitted(m2))

penguins %>%
  ggplot(aes(x = bill_length_mm,
             y = bill_depth_mm,
             color = species)) +
  geom_point() +
  geom_line(aes(y = yhat, 
                group = species,
                color = species),
            lwd = 1.5) +
  theme_bw(base_size = 13)
```

##

```{r}
#| echo: true
m2
```

. . .

**Dummy Variable**

A variable that is 1 if an observation takes a particular level of a categorical variable and 0 otherwise. A categorical variable with $k$ levels can be encoded using $k - 1$ dummy variables.


## Interpreting coefficients

```{r}
countdown::countdown(1, top = 0)
```

\

```{r}
#| echo: true
m2
```

:::poll
Which is the correct interpretation of the coefficient in front of Gentoo?
:::

:::notes
This one assess their ability to use a conditional interpretation of a regression coefficient (controlling for the other variables in the model...)
:::

## Concept Question 4

Consider the following linear regression output where the variable `school` is categorical and the variable `hours_studied` is numerical.

```{r, out.width = "60%"}
#| eval: false
#| echo: true
#| fig-align: center
lm(GPA ~ hours_studied + school, data = edu)
```

\   
\   

| Coefficients     | Estimate |
| ----------- | ----------- |
| `(Intercept)`      | 2.5       |
| `hours_studied`        | .2        |
| `schoolCal`      | 1       |
| `schoolStanford`        | -1      |

. . . 

## Concept Question 4 (cont.)

-   Say I wanted to create a data frame from the original `edu` dataframe which contains the minimum, median, and IQR for `hours_studied` among each school. In order to do this, I make use of `group_by()` followed by `summarize()`. I save this data frame into an object called `GPA_summary`.


:::poll
What are the dimensions of `GPA_summary`?
:::


:::notes
The correct answer in the poll should be 3x3. The three columns are for each of the statistics calculated; the three rows are for the three levels in the school category (there *is* an additional level beyond `"Cal"` and `"Stanford"`, regardless if it is not stated). 

This was a former final question.
:::

```{r}
countdown::countdown(1, top = 0)
```

-->



<!-- ## MLR Geometry -->

<!-- . . . -->

<!-- ```{r} -->
<!-- countdown::countdown(1, top = 0) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- zagat <- read_csv("https://www.dropbox.com/s/c797oanmvdzjegt/zagat.csv?dl=1") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| echo: true -->
<!-- m3 <- lm(price ~ food + decor + geo, data = zagat) -->
<!-- ``` -->

<!-- :::poll -->
<!-- What is the geometry described by this model? -->
<!-- ::: -->

<!-- ::: -->
<!-- The answer is two parallel planes. -->

<!-- This is to test their ability to extrapolate from the examples they've seen in the notes. It's also to have fun spinning a 3D scatter plot. -->

<!-- Students might ask about higher dim, where these models describe "hyperplanes". -->
<!-- ::: -->


<!-- ## -->

<!-- ```{r} -->
<!-- #| code-fold: false -->
<!-- #| echo: false -->
<!-- #| eval: false -->
<!-- library(plotly) -->
<!-- library(reshape2) -->

<!-- grid_points <- 30 -->
<!-- axis_x <- seq(min(zagat$food), -->
<!--               max(zagat$food), -->
<!--               length.out = grid_points) -->
<!-- axis_y <- seq(min(zagat$decor), -->
<!--               max(zagat$decor), -->
<!--               length.out = grid_points) -->
<!-- zagat_plane <- expand.grid(food = axis_x, -->
<!--                          decor = axis_y, -->
<!--                          KEEP.OUT.ATTRS = F) -->
<!-- zagat_plane$price <- predict.lm(m3, newdata = zagat_plane) -->
<!-- z <- acast(zagat_plane, food ~ decor + geo, value.var = "price") -->
<!-- p <- plot_ly(zagat, x = ~food, y = ~decor, z = ~price, showlegend=FALSE) %>% -->
<!--   add_markers(marker = list(size = 5, -->
<!--                             opacity = .6, -->
<!--                             color = "steelblue"), -->
<!--               name = ~restaurant) %>% -->
<!--   config(displayModeBar = FALSE) -->
<!-- p %>% -->
<!--   add_surface(x = axis_x, -->
<!--               y = axis_y, -->
<!--               z = z, showscale=FALSE, -->
<!--               colorscale = list(c(0, 1), c("tan", "tan")), -->
<!--               name = "Linear Model") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- library(plotly) -->
<!-- library(reshape2) -->
<!-- m1 <- lm(price ~ food + decor, data = zagat) -->
<!-- grid_points <- 30 -->
<!-- axis_x <- seq(min(zagat$food),  -->
<!--               max(zagat$food), -->
<!--               length.out = grid_points) -->
<!-- axis_y <- seq(min(zagat$decor),  -->
<!--               max(zagat$decor), -->
<!--               length.out = grid_points) -->
<!-- zagat_plane <- expand.grid(food = axis_x,  -->
<!--                          decor = axis_y,  -->
<!--                          KEEP.OUT.ATTRS = F) -->
<!-- zagat_plane$price <- predict.lm(m1, newdata = zagat_plane) -->
<!-- z <- acast(zagat_plane, food ~ decor, value.var = "price") -->

<!-- m2 <- lm(price ~ food + decor + geo, data = zagat) -->

<!-- zagat_plane_geo <- zagat_plane %>% -->
<!--   mutate(geo = 1) -->
<!-- zagat_plane_geo$price <- predict.lm(m2, newdata = zagat_plane_geo) -->
<!-- z_geo <- acast(zagat_plane_geo, food ~ decor, value.var = "price") -->
<!-- zagat_plane_west <- zagat_plane %>% -->
<!--   mutate(geo = 0) -->
<!-- zagat_plane_west$price <- predict.lm(m2, newdata = zagat_plane_west) -->
<!-- z_west <- acast(zagat_plane_west, food ~ decor, value.var = "price") -->

<!-- plot_ly(zagat, x = ~food, y = ~decor, z = ~price) %>% -->
<!--   add_markers(marker = list(size = 5, -->
<!--                             opacity = .6)) %>% -->
<!--   add_surface(x = ~axis_x,  -->
<!--               y = ~axis_y,  -->
<!--               z = ~z_geo, -->
<!--               showscale = FALSE) %>% -->
<!--   add_surface(x = ~axis_x,  -->
<!--               y = ~axis_y,  -->
<!--               z = ~z_west, -->
<!--               showscale = FALSE) -->
<!-- ``` -->

