---
title: "K-Nearest Neighbors"
subtitle: "Practice"
format: pdf
---

\vspace{-2cm}

The data frame below provides a training data set containing five observations, two numerical predictors, and one categorical response variable.
Use this dataset for the first two questions.
For the second questions use the palmerpenguins dataset.

\begin{center}
\begin{tabular}{ c | c | c | c }
  \hline
  Obs. & $X_1$ & $X_2$ & $Y$ \\
  \hline
  1 & 2 & 210 & Orange \\
  2 & 0 & 90 & Blue \\
  3 & 1 & -23 & Orange \\
  4 & 0 & 180 & Blue \\
  5 & 1.2 & 312 & Orange \\
  \hline
\end{tabular}
\end{center}

#### Question 1

Using K Nearest Neighbors with $K = 2$, based on Euclidean distance, predict the class of a new test observation with $x_1 = -1; \quad x_2 = 255$. It may be helpful to sketch a scatterplot of this data.

\vspace{5cm}

#### Question 2

Does this method incorporate the information of each predictor equally? If so, how? If not, how would you tweak the method so that it does?


#### Question 3

We want to fit a KNN model to predict the penguin species from `bill_length_mm` and `body_mass_g`. First make a plot showing bill length, body mass and species. What is the K=3 prediction for a penguin with a 60 mm bill length and 4000 g mass?

```{r}
#| message: false
library(tidyverse)
library(palmerpenguins)
data("penguins")

# we only need these 3 columns for this exercise
penguins <- penguins %>% 
    select(bill_length_mm, body_mass_g, species)
```



#### Question 4

Does this method incorporate the information in each predictor equally? Is so, how? If not, how would you tweak the method so that it does?

Before answering the first part of this, question look at the following two plots. What's going on here?

```{r}
penguins %>% 
    ggplot(aes(x=bill_length_mm, y=body_mass_g, color=species)) + 
    geom_point()
```


```{r}
penguins %>% 
    ggplot(aes(x=bill_length_mm, y=body_mass_g, color=species)) + 
    geom_point() + 
    lims(x=c(0, 6500), y=c(0, 6500))
```
